%---Fall 2013 Math 3600 Syllabus

\documentclass[12pt]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage{url}

\begin{document}

\title{Euclidean Geometry\\ Math 3600, Fall 2013}

\maketitle

\thispagestyle{empty}

\noindent
\textbf{Course Description:}\\
This course is devoted to classical, planar geometry, including basic properties of triangles, circles, quadrilaterals and other polygons, compass and straightedge constructions and the notion of area. 

A focus of the class is working with axiomatic arguments. Special attention will be given to learning to construct, read, write, present and critique mathematical proofs. 
We will spend time discussing the nature and role of definitions and the process of mathematical work.\\[.1in]

\noindent
\textbf{Basic Information:}\\
\begin{tabular}{ll}
Instructor:  & Theron J Hitchman \\
Office: & 327 Wright Hall\\
Office Hours: & \textbf{MWF 11am - noon.}  To meet at another time, make an appointment. \\
contact info: & email:  \url{theron.hitchman@uni.edu} \phantom{wide}  phone: 319-273-2646 \\
Course Web Page: & http://unigeometry.wordpress.com/   \\
 \phantom{wide} & Check this regularly. 
 Lots of important things are at this page. \\
\end{tabular}
{\ }\\[.1in]

\noindent
\textbf{Materials:}\\
\textbf{Text:} 
\emph{Euclid’s Elements}, Heath translation, Green Lion Press, 2002,  ISBN 1888009-19-5. \\
\textbf{Software:} 
GeoGebra is a free, open-source geometry program. \url{www.geogebra.org.}\\ 
\textbf{Drawing Tools:} 
You may wish to use a compass and straightedge set. 
If you do, I recommend a “modern compass” that may be fixed at a given radius.\\[.1in]

\textbf{Assessment:}
I will be using a “standards based assessment” scheme for this course. 
See the relevant handout for more detailed information. 
Grades in this course are based on the following assessments, with emphasis on the first item:
\begin{itemize}
    \item In-class participation, focusing on finding and presenting arguments
    \item a midterm examination
    \item a week-long, take-home, final examination
\end{itemize}
\textbf{If you feel uncertain about your progress in the course at any time, please contact me.}\\[.1in]

\textbf{Accommodations:}
If you have a disability and require accommodations, please contact the instructor early in the semester so that your learning needs may be appropriately met. 
You will need to provide documentation of your disability to the Student Disability Services (SDS) office, located on the top floor of the Student Health Center, (319) 273-2677.




\end{document}




















